---
title: "Ship propulsion calculations"
author: "Dan Kelley"
date: "2018-07-01^[Document reference: dek20180701]"
output: pdf_document
---

**Abstract.** The propulsion calculation of `whalestrike` is compared with published boat horsepower.

# Estimated propulsive for a lobster boat

The following parameters describe a lobster boat advertised at
https://www.kijiji.ca/v-powerboat-motorboat/city-of-halifax/1979-atkinson-lobster-boat/1364491142?enableSearchNavigationFlag=true

```{r}
length <- 11.73                        # m
beam <- 4.63                           # m
draft <- 1.58                          # m
displacement <- 20.46e3                # m^3
speed <- 0.5144 * 7                    # 7 knots converted to m/s
powerEngine <- 110e3                   # power in W
```

The advertisement says that it runs easily at this speed, powered by a (Detroit
6-71) engine at 1300RPM.  The site
https://web.archive.org/web/20120308222058/http://www.barringtondieselclub.co.za/671DetroitDiesel.html
suggests the Detroit Diesel 6-71 provides 101kW at 1200RPM and 119kW at
1400RPM, so the power may be estimated as 110kW.  This must be multiplied by an
efficiency $\epsilon_e$ for the engine, and another efficiency $\epsilon_p$ for
the propeller. Assuming $\epsilon_e=0.45$ (ref 3) and $\epsilon_p=0.4$ (a guess
based on *much* larger propellers discussed in reference 2), the above suggests that
```{r}
power <-  0.45 * 0.4 * powerEngine
```
or `power=`$`r power`$W is the power applied to the ship. Since power is the product of
speed and force, that the force is
```{r}
F1 <- power / speed
```

Thus, the advertisement suggests a propulsion force of `F1=`$`r round(F1)`$ N.

# Estimation of force from parameterizations

As an estimate of wetted area, take the product of length and girth estimated as
```{r}
Ss <- length * (beam + 2 * draft)
```
or `Ss=`$`r round(Ss, 1)`$m^2.

The default `Cs` in `whalestrike` is presently 0.01, and this yields

```{r}
library(whalestrike)
state <- list(xs=-1.2, vs=0.5144*7, xw=0, vw=0)
parms <- parameters(ms=20000,
              Ss=Ss,
              mw=12330, Sw=63,
              lw=10,
              delta=0.02,
              Es=2e7, theta=45, Eb=6e5, beta=0.2, Cs=0.01, Cw=0.0025)
F2 <- -shipWaterForce(7*0.5144, parms)
```
or `F2=`$`r round(F2)`$N.

Thus, the `whalestrike` application infers a propulsion force of `F1=`$`r round(F2)`$N.

The ratio of force estimates is $F1/F2=`r round(F1/F2, 2)`$.

```{r echo=FALSE}
library(whalestrike)
state <- list(xs=-1.2, vs=0.5144*7, xw=0, vw=0)
parms <- parameters(ms=20000,
              Ss=Ss,
              mw=12330, Sw=63, lw=10,
              delta=0.02,
              Es=2e7, theta=45, Eb=6e5, beta=0.2, Cs=0.01, Cw=0.0025)
F2b <- -shipWaterForce(7*0.5144, parms)
```

**ASSESSMENT:** the above suggests that the ship drag calculated by the
`whalecollision` R-app (which uses `whalestrike` for its calculations) is
reliable within a factor of about 2.  This can be judged a successful result,
since ship fouling can yield factor-of-2 variations in ship drag force (e.g.
fig 1 of ref 4), and since the use in `whalecollision` is only to determine
post-collision drag forces, which are so small that they barely affect
velocities on the sub-second timescales of the simulation.


**References**

1. http://web.mit.edu/13.012/www/handouts/propellers_reading.pdf
2.  https://marine.mandieselturbo.com/docs/librariesprovider6/propeller-aftship/basic-principles-of-propulsion.pdf?sfvrsn=0
(fig 28 gives prop efficiency of 0.51 to 0.56 for VERY big props, decreasing with smaller ones but we likely have 0.5m diameter, way way off scale)
3. https://en.wikipedia.org/wiki/Diesel_engine
4. Woods Hole Oceanographic Institution, 1952. Marine fouling and its prevention. United States Naval Institute, Annapolis MD.

