---
title: Ship propulsion calculations
author: Dan Kelley
date: 2018-06-28
---

**Abstract.** The propulsion calculation of `whalestrike` is compared with published boat horsepower.

# Estimated propulsive for a lobster boat

The following parameters describe a lobster boat advertised at
https://www.kijiji.ca/v-powerboat-motorboat/city-of-halifax/1979-atkinson-lobster-boat/1364491142?enableSearchNavigationFlag=true

```{r}
length <- 11.73                        # m
beam <- 4.63                           # m
draft <- 1.58                          # m
displacement <- 20.46e3                # m^3
speed <- 0.5144 * 7                    # the stated 7 knots is here converted to m/s
powerEngine <- 110e3 # power in W
```

The advertisement says that it runs easily at this speed, powered by a (Detroit
6-71) engine at 1300RPM.  The site
https://web.archive.org/web/20120308222058/http://www.barringtondieselclub.co.za/671DetroitDiesel.html
suggests the Detroit Diesel 6-71 provides 101kW at 1200RPM and 119kW at
1400RPM, so the power may be estimated as 110kW.  This must be multiplied by an
efficiency $\epsilon_e$ for the engine, and another efficiency $\epsilon_p$ for
the propeller. Assuming $\epsilon_e=0.45$ (ref 3) and $\epsilon_p=0.4$ (a guess
based on *much* larger propellers discussed in reference 2), the above suggests that
```{r}
power <-  0.45 * 0.4 * powerEngine
```
or $`r power`$W is the power applied to the ship. Since power is the product of
speed and force, that the force is
```{r}
F1 <- power / speed
```

Thus, the advertisement suggests a propulsion force of `F1=`$`r round(F1)`$ N.

# Estimation of force from parameterizations

As an estimate of wetted area, take the product of length and girth estimated as
```{r}
Ss <- length * (beam + 2 * draft)
```
or `r round(Ss, 1)`m^2.

The default `Cs` in `whalestrike` is presently 0.005, and this yields

```{r}
library(whalestrike)
state <- list(xs=-1.2, vs=0.5144*7, xw=0, vw=0)
parms <- parameters(ms=20000,
              Ss=Ss,
              B=4.6, D=1.6, mw=12330, Sw=63,
              lw=10,
              delta=0.02,
              Es=2e7, theta=45, Eb=6e5, beta=0.2, Cs=0.0025*2, Cw=0.003)
F2 <- -shipWaterForce(7*0.5144, parms)
```
or `r round(F2)`N.

Thus, the `whalestrike` application infers a propulsion force of `F1=`$`r round(F2)`$ N.

The ratio of force estimates is `r round(F1/F2, 1)`.

```{r echo=FALSE}
library(whalestrike)
state <- list(xs=-1.2, vs=0.5144*7, xw=0, vw=0)
Cs <- 0.0025 * 3
parms <- parameters(ms=20000,
              Ss=Ss,
              B=4.6, D=1.6, mw=12330, Sw=63,
              lw=10,
              delta=0.02,
              Es=2e7, theta=45, Eb=6e5, beta=0.2, Cs=0.0025*3, Cw=0.003)
F2b <- -shipWaterForce(7*0.5144, parms)
```
However, `Cs`=`r Cs` gives ratio `r round(F1/F2b, 1)`.


**ASSESSMENT:** the above suggests that the app may be giving reasonable forces.

**References**

* 1. http://web.mit.edu/13.012/www/handouts/propellers_reading.pdf
* 2. https://marine.mandieselturbo.com/docs/librariesprovider6/propeller-aftship/basic-principles-of-propulsion.pdf?sfvrsn=0 (fig 28 gives prop efficiency of 0.51 to 0.56 for VERY big props, decreasing
  with smaller ones but we likely have 0.5m diameter, way way off scale)
* 3. https://en.wikipedia.org/wiki/Diesel_engine
