---
title: Boat parameters
author: Dan Kelley
---

**Abstract.** Some tests on boat properties (especially drag) are carried out.

```{r}
L <- 11.73
B <- 4.63
D <- 1.58
displacement <- 20.46e3
knots <- 7
power <- 110e3
```

A lobster boat advertised for sail at
https://www.kijiji.ca/v-powerboat-motorboat/city-of-halifax/1979-atkinson-lobster-boat/1364491142?enableSearchNavigationFlag=true
lists length `r L`m, beam `r B`m, draft `r D`m, and displacement `r displacement` tonnes.
The easy-running speed is stated as `r knots` knots with the engine (Detroit 6-71) at
1300RPM.

The site
https://web.archive.org/web/20120308222058/http://www.barringtondieselclub.co.za/671DetroitDiesel.html
suggests the Detroit Diesel 6-71 provides 101kW at 1200RPM and 119kW at
1400RPM, so the power is likely 110kW.  This is *not* power into the water,
but, since I don't have any efficiencies handy, I will assume that it is.

Since power is the product of force and speed, we infer force
```{r}
F <- power / (knots * 0.514)
```
or `r round(F)` N.

As an estimate of wetted area, take the product of length and girth estimated as
```{r}
Ss <- L * (2*D * B)
Ss
```

These numbers are set as the default for the collision app (version 6) and it
calculates propulsive force of 1140N. The ratio to the present estimate is thus
`r 2280/F`. Assuming 45 percent efficiency of a diesel engine, the ratio
becomes `r 2280/F/0.45`. Assuming 40 percent propellor efficiency (ref 2 guess) gives
`r 2280/F/0.45/0.4`.

**ASSESSMENT:** the above suggests that the app may be giving reasonable forces.

**TO DO:** (1) get estimates for efficiency of propellor etc, perhaps from. (2) read ref 2, which looks like a great (modern) update on things.

**References**

* 1. http://web.mit.edu/13.012/www/handouts/propellers_reading.pdf
* 2. https://marine.mandieselturbo.com/docs/librariesprovider6/propeller-aftship/basic-principles-of-propulsion.pdf?sfvrsn=0 (fig 28 gives prop efficiency of 0.51 to 0.56 for VERY big props, decreasing
  with smaller ones but we likely have 0.5m diameter, way way off scale)
