---
title: "Using the whalestrike package"
author: "Dan Kelley"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Using the whalestrike package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- 1. edit this file in vignettes/whalestrike.Rmd -->
<!-- 2. run devtools::build_vignettes() -->
<!-- 3. build the package -->


```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the whalestrike
package to simulate the collision of a ship with a whale.

# Introduction

This package solves Newton's second law for a simple model of a ship colliding
with a whale. This is a stripped-down model that does not attempt to simulate
the biomechanical interactions that can be simulated in finite-element
treatments such as that of Raymond (2007).  The goal is to establish a
convenient framework for rapid computation of impacts in a wide variety of
conditions. The model runs quickly enough to keep up with mouse movements to
select conditions, in a companion R shiny app. That app lets the user see the
effects of changing contact area, ship speed, etc., as a way to build intuition
for scenarios ranging from collision with a slow-moving fishing boat to
collision with a thin dagger-board of a much swifter racing sailboat. Another
advantage of the simple formulation is that it makes it easy to modify various
dynamical and biomechanical parameters, to add new forces, and to explore a
range of criteria for whale damage.

The documentation for [strike](../help/strike) provides an example of using the
main functions of this package, and so it is a good place to start. A companion
manuscript is intended to provide more detail about the analysis and the
context.

# GUI usage

The following launches a GUI application that should be somewhat self-explanatory.
```
library(whalestrike)
app()
```

# Non-GUI usage

## Overview plot

The following shows how to run a simulation and produce a simple summary plot.
(Setting the second argument of `plot()` gives access to about a dozen plot
types.) The left panel shows position of ship as a dashed curve, while the
whale is indicated with three curves, the top being an imagined point mass in
the whale interior, the one below it being the interface between sub-layer and
blubber, and the one below that being the whale skin. (Actually, the skin is so
thin that what seems to be a thick line is actually two lines.) The impact
occurs when the ship line contacts the whale-skin line, at about 0.3s, and
lasts about 0.5s.

The middle shows a cross section of whale skin, blubber and sub-layer. Note the
thinning of the bottom two layers, indicative of high compressive stress
associated with the impact of the boat.

The right panel shows the Lethality Index (LI), a quantity that was devised by
examination of the published reports of whale strikes.  In the analysis,
reports of "No Injury" and "Minor Injury" were assessed as having a lethality
of 0, while reports of "Severe Injury" or "Mortality" were assigned as having a
lethality of 1.  These observed lethalities were then fitted with a logistic
model as a function of the base-10 logarithm of the maximum compressive stress
during a simulation of the event using [strike()], using published or inferred
whale and ship characteristics. Simulations in which the peak LI exceeds 0.5,
as in the case shown below, are indicative of serious risk to the life of the
whale.

```{r results="hide", fig.width=7, fig.height=3}
library(whalestrike)
t <- seq(0, 1, length.out=200)
state <- list(xs=-2.5, vs=knot2mps(10), xw=0, vw=0) # 10 knot ship
parms <- parameters() # defaults
sol <- strike(t, state, parms)
par(mfcol=c(1, 3), mar=c(3.3, 3, 1, 2), mgp=c(2, 0.7, 0), cex=0.7)
plot(sol)
```

**Exercises.** Rerun the simulation with different parameters, e.g. (1) lower
the speed until the peak LI is below 0.5, (2) vary whale length to consider the
effects of strikes on immature animals, and (3) model strikes near the
mandible, by altering the thickness, modulus, and strength of the sublayer.
(See the documentation for [parameters](../help/parameters) to learn how to set
the relevant parameters of the ship and whale, and use that for
[strike](../help/strike) to learn about setting vessel speed and the simulation
time interval.)


## Threat dependence on impact area

This example shows how to extract information from a sequence of simulations,
to create a graphical display that is not part of the default list. For
clarity, this is done with a loop rather than using e.g `lapply`.
```{r results="hide", fig.width=7, fig.height=4}
library(whalestrike)
t <- seq(0, 1, length.out=200)
state <- list(xs=-1.5, vs=knot2mps(10), xw=0, vw=0) # 10 knots
area <- seq(0.2, 2, length.out=100)
stress <- rep(NA, length.out=length(area)) # compressive stress [MPa]
for (i in seq_along(area)) {
    L <- sqrt(area[i])
    parms <- parameters(Ly=L, Lz=L)
    sol <- strike(t, state, parms)
    stress[i] <- max(sol$WCF$stress) / 1e6
}
danger <- parms$s[2] / 1e6
plot(area, stress, type="l", xlab="Area [m^2]", ylab="Stress [MPa]",
     ylim=c(0, max(stress)))
lines(area[stress>=danger], stress[stress>=danger], lwd=3)
abline(h=danger, lty="dashed")
mtext(sprintf("Compression stress [MPa]\n(injurious if > %.2f MPa)", danger),
      side=3, line=1)
```

**Exercise.** Look at the help for `plot.strike()`, to learn how to examine
skin stress in the y and z directions separately, and then explore the effect
of adjusting impact geometry, at constant area.

## Threat dependence on blubber thickness and ship speed

This example shows how to create a matrix of simulation results, in order to
display dependence of a result upon two parameters. Here, the display shows the
maximum compressive strain as a function of ship speed and blubber thickness.
Note that the contour lines are thickened when the strain exceeds an estimate
of the ratio of blubber ultimate tensile strength to blubber modulus.

```{r results="hide", fig.width=7, fig.height=4}
library(whalestrike)
t <- seq(0, 1, length.out=200)
## Hint: the following creates x and y of different lengths,
## so that mismatches between row/col and i/j values will
## yield errors.
l2 <- seq(0.1, 0.25, length.out=9) # blubber thickness
speedK <- seq(4, 15, length.out=10) # in knots
speed <- knot2mps(speedK)
## stress = peak stress during each simulation, in MPa
stress <- matrix(NA, nrow=length(speed), ncol=length(l2))
l <- parameters()$l
for (i in seq_along(l2)) {
    for (j in seq_along(speed)) {
        state <- list(xs=-1.5, vs=speed[j], xw=0, vw=0)
        parms <- parameters(l=c(l[1], l2[i], l[3], l[4]))
        sol <- strike(t, state, parms)
        stress[j, i] <- max(sol$WCF$stress) / 1e6
    }
}
danger <- parms$s[2] / 1e6
contour(speedK, l2, stress, levels=seq(0, danger, 0.1),
        xlab="Speed [knots]", ylab="Blubber thickness [m]")
contour(speedK, l2, stress, level=danger, lty=2, add=TRUE, drawlabels=FALSE)
contour(speedK, l2, stress, level=seq(3, danger, -0.1), lwd=2, add=TRUE)
mtext(sprintf("Compression stress [MPa]\n(injurious if > %.2f MPa, dashed contour)", danger),
      side=3, line=1)
```

**Exercise.** Explore the effect of varying sublayer properties and thickness
on this graph.


# Further reading

The documentation for the `whalestrike` package provides many references to the
literature. A good start is to use `help("strike",package="whalestrike")` to
get help on a key function of the package.

