---
title: "Using the whalestrike package"
author: "Dan Kelley"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_caption: yes
    fig_width: 5
    fig_height: 5
    dpi: 72
    dev.args: list(pointsize=11)
vignette: >
  %\VignetteIndexEntry{Using the whalestrike package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<!-- HOW TO BUILD THE VIGNETTE. -->
<!-- 1. edit this file in vignettes/whalestrike.Rmd -->
<!-- 2. run devtools::build_vignettes() -->
<!-- 3. build the package -->


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

**Abstract.** This vignette explains the basics of using the whalestrike
package to simulate the collision of a ship with a whale.

# Introduction

This package solves Newton's second law for a simple model of a ship colliding
with a whale. This is a stripped-down model that does not attempt to simulate
the biomechanical interactions that can be simulated in finite-element
treatments such as that of Raymond (2007).  The goal is to establish a
convenient framework for rapid computation of impacts in a wide variety of
conditions. The model runs quickly enough to keep up with mouse movements to
select conditions, in a companion R shiny app. That app lets the user see the
effects of changing contact area, ship speed, etc., as a way to build intuition
for scenarios ranging from collision with a slow-moving fishing boat to
collision with a thin dagger-board of a much swifter racing sailboat. Another
advantage of the simple formulation is that it makes it easy to modify various
dynamical and biomechanical parameters, to add new forces, and to explore a
range of criteria for whale damage.

The documentation for \code{\link{strike}} provides an example of using the
main functions of this package, and so it is a good place to start. A companion
manuscript is intended to provide more detail about the analysis and the
context.

# Examples

## Overview plot

```{r results="hide"}
library(whalestrike)
t <- seq(0, 0.8, length.out=500)
state <- c(xs=-1.5, vs=5, xw=0, vw=0)

ms <- 20e3                             # ship mass
Ss <- 11.73*(2*1.58+4.63)              # ship wetted area (max)
impactWidth <- 0.5                     # impact width (guess, for pointed bow)
impactHeight <- 1.58                   # impact height (draft)
lw <- 10                               # whale length
mw <- whaleMassFromLength(lw)          # whale mass
Sw <- whaleAreaFromLength(lw,"wetted") # whale area
parms <- parameters(ms=ms, Ss=Ss, impactWidth=impactWidth, impactHeight=impactHeight,
                    lw=lw, mw=mw, Sw=Sw)
sol <- strike(t, state, parms)
par(mfcol=c(3, 3), mar=c(2, 3, 0.5, 0.5), mgp=c(2, 0.7, 0), cex=0.7)
plot(sol)
```

## Acceleration dependence on blubber thickness

```{r results="hide"}
library(whalestrike)
t <- seq(0, 1, length.out=500)
state <- c(xs=-1.5, vs=5, xw=0, vw=0)
beta <- seq(0.1, 0.3, length.out=100)
maxAccel <- rep(NA, length(beta))
for (i in seq_along(beta)) {
    parms <- parameters(ms=ms, Ss=Ss, impactWidth=impactWidth, impactHeight=impactHeight,
                        lw=lw, mw=mw, Sw=Sw,
                        delta=0.02, Es=2e7, theta=45,
                        Eb=1e6, beta=beta[i])
    sol <- strike(t, state, parms)
    maxAccel[i] <- max(abs(diff(sol$vw))) / (t[2] - t[1])
}
plot(beta, maxAccel, type="l", xlab=expression("Blubber thickness [m]"), ylab="Max. Acceleration [m/s^2]")
```

## Blubber-damage dependence on blubber thickness and ship speed

```{r results="hide"}
library(whalestrike)
t <- seq(0, 1, length.out=500)
## Hint: making x and y of different lengths, to avoid row,col
## versus i,j confusion.
beta <- seq(0.1, 0.3, length.out=9)
speedK <- seq(2, 15, length.out=10) # in knots
speed <- 0.5144 * speedK
maxStrain <- matrix(NA, nrow=length(speed), ncol=length(beta))
for (i in seq_along(beta)) {
    for (j in seq_along(speed)) {
        state <- c(xs=-1.5, vs=speed[j], xw=0, vw=0)
        parms <- parameters(ms=ms, Ss=Ss, impactWidth=impactWidth, impactHeight=impactHeight,
                            lw=lw, mw=mw, Sw=Sw,
                            delta=0.02, Es=2e7, theta=45,
                            Eb=1e6, beta=beta[i])
        sol <- strike(t, state, parms)
        maxStrain[j, i] <- max(sol$WCF$strain)
    }
}
danger <- 0.8 / 1.2
contour(speedK, beta, maxStrain, level=seq(0, danger, 0.1),
        xlab="Speed [knots]", ylab="Blubber thickness [m]")
contour(speedK, beta, maxStrain, level=seq(1, danger, -0.1), lwd=3, add=TRUE)
contour(speedK, beta, maxStrain, level=2/3, lwd=3, lty=2, add=TRUE)
## 2/3 = 0.8/1.2 is Grear et al. 2038 result (page 144, left column) that
## mean (seal) blubber strength is 0.8MPa, whereas mean (seal) blubber
## modulus is 1.2MPa.
mtext("Contours of max strain (dangerous if > 2/3)", side=3)
```


# Further reading

* Fortune, S.M.E., Trites, A.W., Perryman, W.L., Moore, M.J., Pettis, H.M.,
  Lynn, M.S., 2012. Growth and rapid early development of North Atlantic right
  whales (Eubalaena glacialis). Journal of Mammalogy 93, 1342–1354.
  https://doi.org/10.1644/11-MAMM-A-297.1

* Grear, M.E., Motley, M.R., Crofts, S.B., Witt, A.E., Summers, A.P., Ditsche,
  P., 2018. Mechanical properties of harbor seal skin and blubber − a test of
  anisotropy. Zoology 126, 137–144. https://doi.org/10.1016/j.zool.2017.11.002

* van Manen, J.D., van Oossanen, P., 1988. Resistance, in: Lewis, E.V. (Ed.),
  Principles of Naval Architecture (Second Revision), Volume II - Resistance,
  Propulsion and Vibration. Society of Naval Architects and Marine Engineers
  (U.S.), Jersey City, NJ, pp. 1–125.

* Miller, C.A., Reeb, D., Best, P.B., Knowlton, A.R., Brown, M.W., Moore, M.J., 2011.
  Blubber thickness in right whales Eubalaena glacialis and Eubalaena
  australis related with reproduction, life history status and prey
  abundance. Mar Ecol Prog Ser 438, 267–283.

* Moore, M.J., Knowlton, A.R., Kraus, S.D., McLellan, W.A., Bonde, R.K., 2005.
  Morphometry, gross morphology and available histopathology in North Atlantic
  right whale (Eubalaena glacialis) mortalities (1970 to 2002). Journal of
  Cetacean Research and Management 6, 199–214.

* Ng, L.J., Volman, V., Gibbons, M.M., Phohomsiri, P., Cui, J., Swenson, D.J.,
  Stuhmiller, J.H., 2017. A mechanistic end-to-end concussion model that
  translates head kinematics to neurologic injury. Frontiers in Neurology 8,
  1–18. https://doi.org/10.3389/fneur.2017.00269

* Raymond, J.J., 2007. Development of a numerical model to predict impact
  forces on a North Atlantic Right Whale during collision with a vessel.
  University of New Hampshire.

